Supplementary Note 1: Metrics to compare two trajectories

### 2018-08-24 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics vignette

---
title: "Supplementary Note 1: Evaluation metrics"
output: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{tabularx}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: DejaVu Sans
---

@@

@@

All metrics described here were implemented within our dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).

@@

The $`r label_metric("edge_flip", "latex")`$ score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012]. We implemented a branch and bound approach for this problem, using several heuristics to make speed up the search:

@@

We compared the three scores for several common topologies `r ref("fig", "topology_scores_overview")`. While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores `r ref("fig", "topology_scores_overview", "b")`, although the $`r label_metric("him", "latex")`$ tends to punish cycles more, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations `r ref("fig", "topology_scores_overview", "c")`. Their main difference is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in `r ref("fig", "topology_scores_overview", "d")`.

@@

Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch `r ref("fig", "clustering_scores_overview")`. These clusters of cells can then be compared using one of the many external cluster evaluation measures [@saelensComprehensiveEvaluationModule2018]. When selecting such a metric, we had two conditions:

@@

Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$. If

```{r, results = "asis"}
add_fig(
  result_file("plot_clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
  "clustering_scores_overview",
  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ .")
)
```

@@

Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets. For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories. These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells. We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection. For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.

The number of cell waypoints has a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$. To select an optimal number of cell waypoints, we generated a large dataset with a complex topology, and plotted the accuracy versus the time to compute $`r label_metric("correlation", "latex")`$. TODO

@@

An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory. If the prediction error for a particular cell is low, the more similar its positions are between the two trajectories.

@@

## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of differentially expressed features/genes along the trajectory

One of the main advantages of studying cellular dynamic processes using single-cell -omics data is that the

>>>>

---
title: "Supplementary Note 1: Metrics to compare two trajectories"
output: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    number_sections: true
header-includes:
  - \usepackage{tabularx}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: DejaVu Sans
---

@@

All metrics described here were implemented within the dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).

@@

@@

The $`r label_metric("edge_flip", "latex")`$ score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012]. We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:

@@

We compared the three scores for several common topologies (`r ref("fig", "topology_scores_overview")`). While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`). The $`r label_metric("him", "latex")`$ tends to punish the detection of cycles, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`). The main difference however is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).

@@

Perhaps one of the simplest ways to calculate the similarity between the cellular positions of two topologies is by mapping each cell to its closest milestone _or_ branch (`r ref("fig", "clustering_scores_overview")`). These clusters of cells can then be compared using one of the many external cluster evaluation measures [@saelensComprehensiveEvaluationModule2018]. When selecting such a metric, we had two conditions:

@@

Based on these requirements, and on the analysis in [@saelensComprehensiveEvaluationModule2018], we chose the $\textrm{F1}$ score between the $\textrm{Recovery}$ and $\textrm{Relevance}$. If $C$ and $C'$ are two cell clusters:

$$
Jaccard(c, c') = \frac{|c \cap c'|}{|c \cup c'|}
$$

$$
\textrm{Recovery} = \frac{1}{|C|} \sum_{c \in C}{\max_{c' \in C'}{\textrm{Jaccard(c, c')}}}
$$
$$
\textrm{Relevance} = \frac{1}{|C'|} \sum_{c' \in C'}{\max_{c \in C}{\textrm{Jaccard(c, c')}}}
$$
$$
\textrm{F1} = \frac{2}{\frac{1}{\textrm{Recovery}} + \frac{1}{\textrm{Relevance}}}
$$

```{r, results = "asis"}
add_fig(
  result_file("clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "clustering_scores_overview",
  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the ${label_metric('F1_branches', 'latex')}$, ${label_metric('F1_edges', 'latex')}$ and ${label_metric('F1_branches', 'latex')}$ ."),
  "",
  width = 12,
  height = 5
)
```

@@

Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets. For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (`r ref("fig", "waypoints_overview", "a")`). These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells. We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection. For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.

To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$. To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $`r label_metric("correlation", "latex")`$ at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`). We found that using cell waypoints does not induce a systematic bias in the $`r label_metric("correlation", "latex")`$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.

```{r, results = "asis"}
add_fig(
  result_file("waypoints_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "waypoints_overview",
  glue::glue("Determination of cell waypoints"),
  glue::glue("a) Illustration of the stratified cell sampling using an example dataset (top). Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom). b) Accuracy versus time to calculate ${label_metric('correlation', 'latex')}$"),
  width = 12, height = 8
)
```

@@

An alternative approach to detect whether the positions of cells are similar between two trajectories, is to use the positions of one trajectory to predict the positions within the other trajectory. If the prediction error for a particular cell is low, the higher the similarity between the cellular positions.

@@

## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of dynamical differentially expressed features/genes.

Perhaps the main advantages of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome. This allows the study of dynamic regulatory networks and gene expression modules. Such analyses rely on a "good-enough" cellular ordering, so that the cellular ordering can be used to identify dynamical differentially expressed genes.

To calculate the $`r label_metric("featureimp_cor", "latex")`$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory. Specifically, we calculated the geodesic distances for each cell to all milestones in the trajectory. We then trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell. We then extracted feature importances using the Mean Decrease in Impurity (`importance = 'impurity'` parameter of the `ranger` function), as illustrated in (`r ref("fig", "featureimp_overview.rds")`). The overall importance of a feature (gene) was then equal to the mean importance over all milestones.

```{r, results = "asis"}
add_fig(
  result_file("featureimp_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "featureimp_overview",
  glue::glue(""),
  glue::glue(""),
  width = 12, height = 6
)
```

Random forest regression has two main hyperparameters. The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate estimates of the feature importance. The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) can all get higher on the ranking.
